\graphicspath{{../07Implications/pics/}}

\chapter{Implications}\label{ch:Implications}

\lettrine[lines=2]{\color{darkocre}I}{n this chapter we'll examine how quantum physics affects our} understanding of the world at the fundamental level. We will analyze two big questions: What are the ultimate "building blocks" of reality, and what we can know about them. 

\begin{myprereq}{Prerequisite Knowledge}
	To fully understand the material of this chapter, readers should be comfortable with the following concepts:
	
	\begin{itemize}
		\item \phantom{phantom}
		\vspace{-0.5cm}
		\item State
		\item Dynamical equations
	\end{itemize}	
\end{myprereq}

\section{Two Ologies}
Ontology and epistemology.

\section{Apriori Values}
Elements of reality.

\section{Complimentarity}
Bohr's ideas.

\section{Phenomena}
Bohr's ideas.

\section{Wholeness}
David Bohm's ideas.

\section{It From Bit}
Wheeler's ideas.

\section{Non-Kolmogorov Probability}

\section*{Chapter Highlights}
{\setstretch{1.5}\chhc
	\it	
	\begin{itemize}
		\item Tensors find application in various areas of science and math.
		\item Geometrical properties of surfaces and spaces can be described
		using metric tensor.
		\item Physical properties of solids are often anisotropic -- depend on
		the direction of applied ``force''. Such properties are best
		described by various tensors: stress tensor, mobility tensor,
		piezoelectric tensor, and others.
		\item At the fundamental level electric and magnetic fields are united
		in a single physical object -- electromagnetic field. Electromagnetic
		field is described by an antisymmetric tensor of the second rank.
	\end{itemize}
	
}